Variance reduction for MCMC algorithms via martingale representations

### 2020-05-11 aap/Revision/Response_1.tex

We thank the referee for carefully reading our manuscript and for pertinent
comments which helped us to improve the paper. In the new version of our
paper, we have taken all comments and suggestion into account.  Below are some additional responses to the
comments by Referee 1

\section*{Main remarks}

\begin{itemize}

\item \textit{... The authors make an attempt to quantify the variance reduction obtained by the proposed approach, as well as the computational complexity to achieve this variance reduction. On these topics several results are presented as Theorems and Propositions  ... I would recommend the authors to phrase this results with clear conditions in such a way that it becomes directly accessible to the reader ... where the bound is phrased in an easily readible way involving constants that are defined in the statement of the theorem and which should not depend on infinite series and conditional expectations.}
\par
Let us first stress that  current literature on MCMC  mainly focuses on  convergence of MCMC algorithms in Total Variation and Wasserstein distances thus  ignoring a sample error of these algorithms. Only few references studied MSE error of the Langevin algorithm (see e.g.  Durmus and  Moulines (2017)). In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms. The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  the cost-to-variance (product of cost and variance) criterion. Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest. Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation. In  Section~4 we  also explain in an accessible way why our approach leads to  smaller cost-to-variance characteristics. These results are based on a thorough analysis of the variance reduced estimate postponed to Section 5. In the revised version we significantly weaken  assumptions  and prove our main result (Theorem~10 and Corollary~11) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).   Note that all assumptions needed are now collected in Theorem~10 and Corollary~11. For convenience of the reader we also sketch the main steps of the proof right after 
Theorem~10. We decided to present our main theoretical result in the form of bounds for the functions $A_{q,k}$ as these functions play a crucial role for deriving  the variance bound (32). 
\item \textit{... It is good to see that the technique has beneficial effects on the variance, but it does not become clear what was the computational effort required to achieve this benefit. The reader is left wondering whether or not the same variance reduction could have been achieved simply by drawing more MCMC samples ...}
\par
In the revised version (Section 6.1) we measure the gain of our VR approach by comparing its cost-to-variance characteristic (product of the obtained variance and its cost) to one of the plain MCMC method. This characteristic quantifies  the resulting variance reduction effect relative to computational effort spent. 
\item \textit{... here the setting \(\gamma_k = c k^{-\alpha}, \alpha\in  (0, 1)\) is considered, and presented as  ``the most interesting case''. For many practitioners, $\alpha=0$ would be rather natural. What can be said in this important case? This is also applies to Section 6 ...}
\par
In the revised version we decided to switch to  a constant time step and present all our bounds for this case. On the one hand, this allows us to formulate our main results in a simpler form depending only on a single parameter $\gamma.$ On the other hand, this is indeed the most relevant choice for practitioners.  
\item \textit{ ... here the functions g are assumed uniformly bounded, whereas earlier (end of Section 5) elements in $\Psi$ were proposed to be polynomials. In particular the bound B would be equal to infinity. How can this be reconciled? ...}
\par
In the revised version we skip this result because of size restrictions. 
\end{itemize}

### 2020-05-11 aap/Revision/arxiv/vr-mcmc-arxiv.tex
arxiv update

\documentclass[bj]{imsart}
\makeatletter
\def\set@curr@file#1{%
  \begingroup
    \escapechar\m@ne
    \xdef\@curr@file{\expandafter\string\csname #1\endcsname}%
  \endgroup
}
\def\quote@name#1{"\quote@@name#1\@gobble""}
\def\quote@@name#1"{#1\quote@@name}
\def\unquote@name#1{\quote@@name#1\@gobble"}
\makeatother
\usepackage{graphics}

@@

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
\label{fig:subfig2}}
\caption{Cost-to-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  a two-dimensional standard Gaussian distribution and different test functions.
\label{fig:gaus_cost}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
\label{fig:subfig2}}
\caption{Cost-to-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for the mixture \eqref{eq:gmm} of two-dimensional Gaussian distributions and two  test functions.
\label{fig:gmm_cost}}
\end{figure}

@@

 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  obtained by the variance reduced ULA algorithm for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our method of martingale control variates (MDCV), zero variance estimator (ZV) and the control variates approach based on the empirical spectral variance minimisation (ESVM)
\label{fig:GMM}}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../pictures/banana_shape.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Banana shape density. The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
\label{fig:banana_shape}}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../pictures/logreg_1st_order.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression. The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
 \label{fig:2}}
\end{center}
\end{figure}
In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
\begin{eqnarray*}
\pi(\theta | \mathsf{Y}, \mathsf{X}) \propto \exp \left\{ \mathsf{Y}^T \mathsf{X} \theta - \sum\limits_{i=1}^m \ln{\left(1 + \rme^{\mathsf{X}_i \theta}\right)} - \frac{1}{2\sigma^2} \left\|\theta\right\|_2^2 \right\},
\end{eqnarray*}
Thus we obtain
\begin{eqnarray*}
U(\theta) &=& -\mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{\theta^T \mathsf{X}_i}\right)} + \frac{1}{2\sigma^2} \left\| \theta\right\|_2^2,
\\
\nabla U(\theta) &=& -\mathsf{Y}^T \mathsf{X} + \sum_{i=1}^m \frac{\mathsf{X}_i}{1 + \rme^{-\mathsf{\theta}^T \mathsf{X}_i}} + \frac{1}{\sigma^2} \theta.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we used the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories. We put $n_{trunc} = 50$ and $K = 1$. The target function is taken to be $f(\theta) = \sum_{i=1}^{d}\theta_i$.  In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$. In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use the first order polynomials for the same reasons as in the Gaussian mixtures example.

>>>>

\documentclass[preprint]{imsart}
\makeatletter
\def\set@curr@file#1{%
  \begingroup
    \escapechar\m@ne
    \xdef\@curr@file{\expandafter\string\csname #1\endcsname}%
  \endgroup
}
\def\quote@name#1{"\quote@@name#1\@gobble""}
\def\quote@@name#1"{#1\quote@@name}
\def\unquote@name#1{\quote@@name#1\@gobble"}
\makeatother
\usepackage{graphics}

@@

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{vr_cost_linear_gaussian.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{vr_cost_quadratic_gaussian.png}
\label{fig:subfig2}}
\caption{Cost-to-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  a two-dimensional standard Gaussian distribution and different test functions.
\label{fig:gaus_cost}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{vr_cost_linear_gmm.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{vr_cost_quadratic_gmm.png}
\label{fig:subfig2}}
\caption{Cost-to-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for the mixture \eqref{eq:gmm} of two-dimensional Gaussian distributions and two  test functions.
\label{fig:gmm_cost}}
\end{figure}

@@

 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  obtained by the variance reduced ULA algorithm for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our method of martingale control variates (MDCV), zero variance estimator (ZV) and the control variates approach based on the empirical spectral variance minimisation (ESVM)
\label{fig:GMM}}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{banana_shape.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Banana shape density. The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
\label{fig:banana_shape}}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{logreg_1st_order.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression. The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
 \label{fig:2}}
\end{center}
\end{figure}
In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
\begin{eqnarray*}
\pi(\theta | \mathsf{Y}, \mathsf{X}) \propto \exp \left\{ \mathsf{Y}^T \mathsf{X} \theta - \sum\limits_{i=1}^m \ln{\left(1 + \rme^{\mathsf{X}_i \theta}\right)} - \frac{1}{2\sigma^2} \left\|\theta\right\|_2^2 \right\},
\end{eqnarray*}
Thus we obtain
\begin{eqnarray*}
U(\theta) &=& -\mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{\theta^T \mathsf{X}_i}\right)} + \frac{1}{2\sigma^2} \left\| \theta\right\|_2^2,
\\
\nabla U(\theta) &=& -\mathsf{Y}^T \mathsf{X} + \sum_{i=1}^m \frac{\mathsf{X}_i}{1 + \rme^{-\mathsf{\theta}^T \mathsf{X}_i}} + \frac{1}{\sigma^2} \theta.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we used the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories. We put $n_{trunc} = 50$ and $K = 1$. The target function is taken to be $f(\theta) = \sum_{i=1}^{d}\theta_i$.  In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$. In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use the first order polynomials for the same reasons as in the Gaussian mixtures example.
